An event-sourced and distributed database with IPLD and materialized views

### 2019-09-20 draft.tex
all figures updated, new tables, lots of linkages
Signed-off-by: andrew <andrew@textile.io>

@@

@@

A key principal of ES and append only logs is that all changes to application state are stored as a sequence of events. Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}. The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see section  \ref{sec:ViewsProjections}). In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.

@@

IPLD is used to publish linked data (subject-predicate-object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers. To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{For reference, the Bitcoin Blockchain is a DAG, so is a Git version history.}. This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense \cite{brendanobrienDeterministicQueryingDistributed2017}. In practice, IPLD is represented as objects, each with \texttt{Data} and \texttt{Links} fields, where \texttt{Data} can be a small blob of unstructured, arbitrary binary data, and \texttt{Links} is an array of links to other IPLD objects.

@@

@@

On the downside, and as seen above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots):

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{Divergent heads in a multi-writer Merkle-DAG }
  \label{fig:merkledag}
\end{figure}

@@

@@

All filesystems and databases have some notion of "access control." Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects [ref]. An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object.

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{Example Access Control List.}
  \label{fig:boat1}
\end{figure}

@@

We propose Threads, a protocol and decentralized database that runs on IPFS meant to help decouple apps from user-data. The Threads protocol is designed based on an underlying, CQRS-inspired event-sourcing system for synchronizing data across collaborating peers on a network. Threads offer data ownership and a multi-role data access architecture where owners can set independent permissions for writing, reading, and following data updates. Threads differ from previous solutions by extending on the multiaddress addressing scheme to allow PULL based replica synchronization in addition to more common PUSH based synchronization seen in decentralized protocols. The flexible event-based structure enables client applications to model advanced states through aggregates, views, and custom CRDTs.

@@

In multi-writer systems, determining causal order in all replicas is particularity challenging. As previously discussed (section 2.2.4), a solution based on a Merkle-Clock is only partially ordered. That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy [ref]:

@@

Our solution to deal with divergent Merkle-Clocks is to institute a single-writer rule: A log can only be updated by a single replica or identity. An Event Log is a single-writer Merkle-Clock that can be totally ordered. Separate Event Logs can be composed into advanced structures, including CRDTs [Ref Enes].

Practically speaking, for any given Log, Events are authored by a single IPFS Peer, or Writer.

@@

As seen above in section 2.3.4., a Merkle-Clock is simply a Merkle-DAG of Events:

@@

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{A single-writer Merkle-Clock (Event Log).}
  \label{fig:boat1}
\end{figure}

@@

\begin{example*}
\centering
\begin{minipage}{0.7\textwidth}
\begin{lstlisting}[language=json,firstnumber=1]
# Log multiaddress
 /ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab

@@

Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log

TODO: ref to embed

In practice, Writers are solely responsible for announcing their Log's addresses. This ensures a conflict-free address list without additional complexity. Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers. This allows them to pull (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable.

@@

\begin{figure*}
\centering
\begin{minipage}{0.6\textwidth}
  \includegraphics[width=\linewidth]{figures/Event_Log_With_Encryption.png}
  \caption{The three layers of Log Event encryption.}
  \label{fig:LogEncryption}
  \end{minipage}
\end{figure*}

@@

Textile Logs are designed to be shared, composed, and layered into datasets. Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.

@@

A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization. TIDs are defined with the following format:

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{The thread id formula.}
  \label{fig:boat1}
\end{figure}

@@

See Example \ref{ex:Identity}).
\end{definition}

@@

See Example \ref{ex:Identity}).
\end{definition}

\begin{example*}
\centering
\begin{minipage}{0.6\textwidth}
\begin{lstlisting}[language=json,firstnumber=1]
# Raw identity. V1, 128 bit
bafkxd5bjgi6k4zivuoyxo4ua4mzyy

# ACL enabled identity. V1, 256 bit.
bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya
\end{lstlisting}
\end{minipage}
  \caption{Identity variants.}
  \label{ex:Identity}
\end{example*}

@@

\begin{enumerate}
\item \label{Perf1}Log Writers can receive Events directly from the author.
\item \label{Perf2} Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
    a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events.
\item \label{Perf3} Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
\end{enumerate}

<insert diagram of pulling from a replicating Follower)

@@

The notion of the Follow key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial. This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries. This type of Log replication can act as a data backup mechanism. It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see section  \ref{sec:LogSync}).

@@

\textbf{CRDTs} \\

@@

One of the most important properties of a shared data model is the ability to apply access control rules. There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs. Thread-level access control lists (ACLs) allow creators to specify who can follow, read, write, and delete Thread data. Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition XXX) basis. Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see section 3.3.2 for details). ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible). First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers. Second, because ACLs are JSON documents, ACLs can specify their own editing rules (i.e. allowing multiple Thread participants to modify the ACL) in a self-referencing way.

@@

\begin{example}
\centering
\begin{lstlisting}[language=json,firstnumber=1]
# UUID
bafykrq5i25vd64ghamtgus6lue74k
\end{lstlisting}
  \caption{Sequence ID.}
  \label{ex:SequenceID}
\end{example}

@@

\subsubsection{Textile: The Thread \& Bot Network}

@@

\begin{example*}

@@

    // Key is an optional single-use symmetric key
    // used to encrypt Body.
    Key() []byte
}
\end{lstlisting}
  \caption{This is a full width code block.}

>>>>

\usepackage{dblfloatfix}    % To enable figures at the bottom of page

% trying to keep figures in the right sections instead of pushed to the end
\usepackage[section]{placeins}

% To create tables + create checkmarks
\usepackage{booktabs}
\usepackage{colortbl}
\usepackage{multirow}% http://ctan.org/pkg/multirow

\usepackage{array}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}

\newcommand\Tstrut{\rule{0pt}{2.6ex}}       % "top" strut
\newcommand\Bstrut{\rule[-0.9ex]{0pt}{0pt}} % "bottom" strut
\newcommand{\TBstrut}{\Tstrut\Bstrut} % top&bottom struts

% column centering special type
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
% End table requirements

% Thread ID components fig requrements
\usepackage{stackengine}
% Thread ID equation setup 
\def\textscrm#1{\textrm{\scriptsize#1}}
\stackMath
\def\rldp{1.3ex}
\def\rlht{.8ex}
\def\rlwd{.8pt}
\setstackgap{L}{\rldp}
\def\uvbarR#1{%
  \def\stackalignment{r}\def\stacktype{S}\stackunder[-\rlwd]{%
    \def\stackalignment{c}\def\stacktype{L}\stackunder{#1}{\rule{\rlwd}{\rlht}}%
  }{\setbox0\hbox{$#1$}\rule{.5\wd0}{\rlwd}}%
}
\def\uvbarL#1{%
  \def\stackalignment{l}\def\stacktype{S}\stackunder[-\rlwd]{%
    \def\stackalignment{c}\def\stacktype{L}\stackunder{#1}{\rule{\rlwd}{\rlht}}%
  }{\setbox0\hbox{$#1$}\rule{.5\wd0}{\rlwd}}%
}
\def\uvbar#1#2{%
  \def\stacktype{S}\def\stackalignment{c}\def\useanchorwidth{T}\stackunder[0pt]{%
    \def\stacktype{L}\setbox0\hbox{${}#1{}$}%
    \stackunder[\rldp]{{}#1{}}{\rule{\wd0}{\rlwd}}%
  }{\def\stacktype{S}\stackunder[1pt]{\rule{\rlwd}{\rlht}}{\scriptstyle#2}}%
}
\newcommand\ubar[2]{\uvbarR{\textrm{}}\uvbar{\textrm{#1}}{\textscrm{#2}}%
  \uvbarL{\textrm{}}}
\newcommand\phubar[2]{\uvbarR{\textrm{}}\uvbar{\phantom{\textrm{#1}}}{\textscrm{#2}}%
  \uvbarL{\textrm{}}}

% End thread Id requirements

@@

% bolds "Example X" to match Figure X
\usepackage[labelfont=bf]{caption}
\DeclareCaptionType{example}

@@

A key principal of ES and append only logs is that all changes to application state are stored as a sequence of events. Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}. The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see Section  \ref{sec:ViewsProjections}). In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.

@@

IPLD is used to publish linked data (subject-predicate-object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers. To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{Other examples of Dags include the Bitcoin Blockchain or a Git version history.}. This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense \cite{brendanobrienDeterministicQueryingDistributed2017}. In practice, IPLD is represented as objects, each with \texttt{Data} and \texttt{Links} fields, where \texttt{Data} can be a small blob of unstructured, arbitrary binary data, and \texttt{Links} is an array of links to other IPLD objects.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/Divergent_Heads.png}
  \caption{Divergent heads in a multi-writer Merkle-DAG }
  \label{fig:merkledag}
\end{figure}

@@

On the downside, and as discussed above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots) (see Figure \ref{fig:merkledag})

@@

% blocks the merkledag image from being rendered past here
\FloatBarrier

@@

All filesystems and databases have some notion of "access control." Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects [ref]. An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object (see Figure /ref{tbl:ACL}).

\begin{table*}
\caption{Example Access Control List.}
    \centering
    \begin{tabular} [t]{l>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering\arraybackslash}p{0.2\linewidth}}
\toprule
&Create&Delete&Edit&Read\\
\midrule
John&\checkmark&\checkmark&\checkmark&\checkmark\\
Jane&-&-&-&\checkmark\\
Mary&-&-&\checkmark&\checkmark\\
\bottomrule
    \end{tabular}
    \label{tbl:ACL}
\end{table*}%

@@

We propose Threads, a protocol and decentralized database that runs on IPFS meant to help decouple apps from user-data. The Threads protocol is designed based on an underlying, CQRS-inspired event-sourcing system for synchronizing data across collaborating peers on a network. Threads offer data ownership and a multi-role data access architecture where owners can set independent permissions for writing, reading, and following data updates. Threads differ from previous solutions by extending on the multiaddress addressing scheme to allow pull-based replica synchronization in addition to more common push-based synchronization seen in decentralized protocols. The flexible event-based structure enables client applications to model advanced states through aggregates, views, and custom CRDTs.

@@

% [!b] indicates that it will try to be on the bottom of the page
\begin{figure}[!b]
  \includegraphics[width=\linewidth]{figures/Event_Log.png}
  \caption{A single-writer Merkle-Clock (Event Log).}
  \label{fig:EventLog}
\end{figure}

In multi-writer systems, determining causal order in all replicas is particularity challenging. As previously discussed (section 2.2.4), a solution based on a Merkle-Clock is only partially ordered. That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy [ref].

@@

Our solution to deal with divergent Merkle-Clocks is to institute a single-writer rule: A log can only be updated by a single replica or identity. An Event Log is a single-writer Merkle-Clock that can be totally ordered (Figure \ref{fig:EventLog}). Separate Event Logs can be composed into advanced structures, including CRDTs [Ref Enes].

For any given Log, Events are authored by a single IPFS Peer, or Writer.

@@

As seen above in Section \ref{sec:merkleclocks}., a Merkle-Clock is simply a Merkle-DAG of Events:

@@

% blocks the EventLog image from being rendered past here
\FloatBarrier

\begin{example*}[!b]
\centering
\begin{minipage}{0.7\textwidth}
\begin{lstlisting}[language=json,firstnumber=1]
# Log multiaddress
 /ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab

@@

@@

Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log (Example {ex:Multiaddress}).

In practice, Writers are solely responsible for announcing their Log's addresses. This ensures a conflict-free address list without additional complexity. Some Followers may be in the business of replicating Logs (see Section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers. This allows them to pull (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).

@@

\begin{figure*}[!t]
\centering
\begin{minipage}{0.6\textwidth}
  \includegraphics[width=\linewidth]{figures/Event_Log_With_Encryption.png}
  \caption{The three layers of Log Event encryption.}
  \label{fig:LogEncryption}
  \end{minipage}
\end{figure*}

@@

Textile Logs are designed to be shared, composed, and layered into datasets (Figure \ref{fig:LogEncryption}). Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.

@@

A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization. TIDs are defined with the following format {see Figure \ref{fig:ThreadID}

% [!b] bottom page
\begin{figure*}[!b]
\centering
\caption{Components of a Thread Identity.}
\par\vspace{.9\baselineskip}\leavevmode%
\par{Thread ID = 0x62\quad}\ubar{0x01}{Version}\quad0x55\quad\ubar{0x539bc1dc03ee8cb5d478e41cc8a4546e}{Random Number} 
\par\vspace{-.6\baselineskip}\leavevmode%
\phantom{Thread ID = }\phubar{0x62}{Multibase}%
  \phantom{\quad0x01\quad}\phubar{0x55}{Variant} \phantom{\quad7513b4dcf4efa291ffa92221659f19e2}
\par\vspace{-.6\baselineskip}\leavevmode
\label{fig:ThreadID}
\end{figure*}

@@

See Example \ref{ex:Identity}a).
\end{definition}

@@

See Example \ref{ex:Identity}b).
\end{definition}

\begin{example*}
\centering
\begin{minipage}{0.6\textwidth}
\begin{lstlisting}[language=json,firstnumber=1]
# (a) Raw identity. V1, 128 bit
bafkxd5bjgi6k4zivuoyxo4ua4mzyy

# (b) ACL enabled identity. V1, 256 bit.
bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya
\end{lstlisting}
\end{minipage}
  \caption{Identity variants.}
  \label{ex:Identity}
\end{example*}

@@

\begin{enumerate}
\item \label{Perf1}Log Writers can receive Events directly from the author.
\item \label{Perf2} Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
    a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
\item \label{Perf3} Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
\end{enumerate}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/Pulling.png}
  \caption{A pull-based request from a Follower.}
  \label{fig:Pulling}
\end{figure}

@@

The notion of the Follow key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial. This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries. This type of Log replication can act as a data backup mechanism. It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see Section  \ref{sec:LogSync}).

@@

\textbf{CRDTs} \label{sec:TexCRDT} \\

@@

One of the most important properties of a shared data model is the ability to apply access control rules. There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs. Thread-level access control lists (ACLs) allow creators to specify who can follow, read, write, and delete Thread data. Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition XXX) basis. Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see Section \ref{sec:TexCRDT}). ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible). First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers. Second, because ACLs are JSON documents, ACLs can specify their own editing rules (i.e. allowing multiple Thread participants to modify the ACL) in a self-referencing way.

@@

% [!b] moves it to the bottom of the page
\begin{example}[!b]
\centering
\begin{lstlisting}[language=json,firstnumber=1]
# UUID
bafykrq5i25vd64ghamtgus6lue74k
\end{lstlisting}
  \caption{Sequence ID.}
  \label{ex:SequenceID}
\end{example}

@@

\subsubsection{Textile: The Thread \& Bot Network} \label{sec:Bots}

@@

% [!p] forces it to just take the whole page
\begin{example*}[!p]

@@

    // Key is an optional single-use symmetric key
    // used to encrypt Body.
    Key() []byte
}
\end{lstlisting}
  \caption{Interfaces for an event body, header and node.}
